---
title: "MLB Winning Percentages - A General Inquiry"
author: "brian higginbotham"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, error = FALSE)
```

What is the difference between being the best and worst team in Major League Baseball? This quick dive into the last 100 seasons of current MLB franchises will look at almost 3,000 winning percentages and map their distribution. From the resulting histogram and quantile calculations, we'll get a good sense of the range between a first-place and last-place season.

**Packages**
```{r packages}
# packages used
library(readr)
library(tidyverse)
```

## Retrieving the Data

The team records for each season are from [Baseball Reference](https://www.baseball-reference.com/teams/). Each teams complete season records were exported as ```.csv``` file. (Technically, the files are ```.txt```, but uses the coma as the separating value, so it behaves like a ```.csv```)

We'll need to combine each team's records into one dataframe. Before we do that, let's set up some simple data read and processing commands to ensure we're extracting the information we want from the team files

```{r read and process}
# each team's file is labeled with the team abbreviation. So here, we're reading the Chicago Cubs.
w_perc <- read_csv("../db/mlb/chc.txt")
# renaming the columns that will be kept in the dataframe
w_perc <- rename(w_perc, year = Year, perc = 'W-L%', team=Tm)
# selecting only the renamed columns for the dataframe
w_perc <- select(w_perc, year, team, perc)
# filtering the dataframe to the last 100 years
w_perc <- filter(w_perc, year >= 1925, year < 2025)

head(w_perc)
```

## Data Reading & Processing Function

Now that we have the commands to read and process the data into the desired format, we simplify all these separate commands into one function:

```{r process function}
# the argument for read_csv() must be a string format (" "), so we have to enter the function argument, team, as a string
perc_win <- function(team = "str"){
  txtfile <- paste0("../db/mlb/", team,".txt")
  sheet <- read_csv(txtfile)
  sheet <- sheet %>%
    rename(year = Year, perc = 'W-L%', team=Tm) %>%
    select(year, team, perc) %>%
    filter(year >= 1925, year < 2025)
  return(sheet)
}
```

Now let's test our function

```{r run func}
perc_win("chw")
```

Ooof, 2024 was not a good year for the White Sox. But on the brighter side, it does look like our function works correctly - producing 100 season winning percentages for the Chi Town White Sox.

But just how bad were the White Sox compared to all teams over the last 100 seasons?

To find out, we'll need to build a single dataframe from all of the teams datasets and look at the distribution of the winning percentages.

## Combining Data Sets

To combine the data sets into a single dataframe, we'll write a simple ```for loop``` to iterate over the file names utilizing our ```perc_win()``` function and then use ```bind_rows()``` to append each data set to the single data frame.

First, we'll need a list of teams/filenames to iterate over.
```{r filnames}
teams <- list("ana", "ari", "atl", "bal", "bos", "chc", "chw", "cin", "cle", "col", "det", "hou", "kcr", "lad", "mia", "mil", "min", "nym", "nyy", "oak", "phi", "pit", "sdp", "sea", "sfg", "stl", "tbd", "tex", "tor", "wsn")
```

Next, we'll initiate an empty dataframe that the resulting datasets from the ```for loop``` can be added to. Then we initiate the ```for loop``` using the ```perc_win()``` and ```bind_rows()``` functions.
```{r loop}
win_perc <- data.frame()

for(i in teams){
  data = perc_win(i)
  win_perc <- bind_rows(win_perc, data)
}

str(win_perc)
```

Great! The resulting dataframe has over 2200 entries. It looks like we have our winning percentages for each team over the past 100 seasons. (If you're thinking "shouldn't there be 3000 entries since there are 30 teams and 100 seasons?", you should remember that some current franchises did not exist in 1925 and so we get less than 100 entries for those franchises)

## Histogram & Other Explorations

Now let's take a look a histogram of the winning percentages
```{r hist, warning=FALSE}
g <- ggplot(win_perc, aes(x=perc))
g + geom_histogram(bins=100, color="black", fill="red") + xlim(0, 1)
```

Over the last 100 MLB seasons, the range of winning percentages is from about 0.25 to about 0.700. Wow, so it looks like the Chicago White Sox had one of, if not the worst winning record of the last 100 years. Let's confirm who the owner of the worst record really is.
```{r min}
filter(win_perc, perc == min(win_perc$perc))
```

Not quite dead last CHW, but probably a little too close for comfort!

On a lighter note, let's see who owns the largest winning percentage of the past 100 years
```{r max}
filter(win_perc, perc == max(win_perc$perc))
```
Impressive! Too bad they lost 4-0 in that year's World Series.

Let's leave off the outliers and see where 95% of the winning percentages lie
```{r 95 interval}
quantile(win_perc$perc, probs = c(0.025, 0.975))
```
Interesting, it seems that the vast majority (95%) of winning percentages lie between about 33% and 66%. I rounded to these numbers to get a different perspective on this difference - it would appear that last place teams win about 1 out of 3 games while first place teams win about 2 out of 3. Given that the majority of season series are 3 games, this result totally makes sense!